---
title: "Generating embedding/trajectory objects for using in Escort R/Shiny app"
author: "Xiaoru Dong"
output: html_document
fontsize: 13pt
vignette: >
  %\VignetteIndexEntry{Escort}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

This tutorial demonstrates how to create an object that stores embedding and trajectory information. This object can then be used in Shiny apps for advanced analysis steps.

If you have not yet installed Escort, please revisit the main ["Running Escort in R"](https://www.rhondabacher.com/docs-escort/vignettes.html) vignette

```{r, eval=FALSE}
library(Escort)
```


We will create the Shiny app-ready object using the function `prepTraj` provided in the Escort package. It requires three arguments: a reduced dimension matrix, estimated pseudotime, and the fitted trajectory.

We'll use the sample dataset provided in the package to start with.

```{r, eval=FALSE}
data("step0_clean_dyntoy_L3")
```

We will generate an embedding that consists of using 2000 highly variable genes, using PCA as the dimension reduction, and assume we are using Slingshot for the trajectory.

```{r, eval=FALSE}
# get highly variable genes
gene.var <- quick_model_gene_var(norm_counts)
genes.HVGs <- rownames(gene.var)[1:2000]
# use PCA to reduce the dimensionality
embedding1 <- getDR_2D(norm_counts[genes.HVGs,], "PCA")
```

The object `embedding1` represents the `dimred` argument in the `prepTraj` function. If using alternative procedures, make sure that your `dimred` object is a data frame with two columns representing the 2D embedding coordinates, with each row representing a cell. The row names of `dimred` should be unique and accurately reflect cell identifiers.

For Step 3, we also need to go ahead and fit a trajectory to ensure the embedding works well not just independently, but also in the context of a specific method. Here we will proceed with Slingshot, however, other methods can be utilized.

```{r, eval=FALSE}
library(slingshot)
library(mclust)
cls1 <- Mclust(embedding1)$classification
ti_out1 <- slingshot(data=embedding1, clusterLabels=cls1)
rawpse1 <- slingPseudotime(ti_out1, na=T)
ls_fitLine1 <- lapply(slingCurves(ti_out1), function(x) x$s[x$ord,])
```

The object `rawpse1` will be input to the `PT` argument in the `prepTraj` function. Note that this object should be either a data frame or a vector. In the case of a data frame, each column should represent a lineage with pseudotime values, and rows should align with cell names in `dimred` object.

The object `ls_fitLine1` will be input to the `fitLine` argument in the `prepTraj` function. Note that this object represents the trajectory with line segments between pairs of points. It is formatted as a data frame, where columns "x0", "y0", "x1", and "y1" denote the starting and ending coordinates of each line segment.


### Prepare the trajectory object:
```{r, eval=FALSE}
embed_obj <- prepTraj(dimred=embedding1, PT=rawpse1, fitLine=ls_fitLine1)
filepath <- "path/to/your_file.rds" # Specify your file path here.
saveRDS(embed_obj, file = filepath)
```

This .rds file is now ready to be loaded in Escort Shiny app for further analysis in Steps 2 and 3!

