---
title: "Generating an Object for Embedding and Trajectory"
author: "Xiaoru Dong"
date: "2024-02-27"
output: html_document
fontsize: 13pt
vignette: >
  %\VignetteIndexEntry{Escort}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

This tutorial demonstrates how to create an object that stores embedding and trajectory information. This object can then be used in Shiny apps for advanced analysis steps.

```{r, eval=FALSE}
library(Escort)
```

### Dimension Reduction Data (`dimred`)
Ensure `dimred` is a dataframe with two columns representing the 2D embedding coordinates. Each row should represent a cell. The row names of `dimred` should be unique and accurately reflect cell identifiers.

```{r, eval=FALSE}
rownames(dimred) <- # Unique cell names.
```

### Pseudotime Estimations (`rawpse`)
`rawpse` should be either a dataframe or a vector. If it's a dataframe, each column should represent a lineage with pseudotime values, and rows should align with cell names in dimred.

Validate that row names in rawpse match those in dimred to ensure alignment.
```{r, eval=FALSE}
rownames(rawpse) <- # Ensure this matches dimred's rownames.
```

### Fitted Curve Line (`fitLine`)
`fitLine` should represent the trajectory with line segments between pairs of points. It is formatted as a dataframe, where columns "x0", "y0", "x1", and "y1" denote the starting and ending coordinates of each line segment.
  
Example data for `fitLine` - adjust to fit your trajectory data
```{r, eval=FALSE}
fitLine <- data.frame(
  x0 = c(1, 2, 3),  # Starting x-coordinates
  y0 = c(1, 2, 3),  # Starting y-coordinates
  x1 = c(2, 3, 4),  # Ending x-coordinates
  y1 = c(2, 3, 4)   # Ending y-coordinates
)
```

### Prepare trajectory object.
```{r, eval=FALSE}
eval_obj <- prepTraj(dimred, PT=rawpse, fitLine=fitLine)
filepath <- "path/to/your_file.rds" # Specify your file path here.
saveRDS(eval_obj, file = filepath)
```

This .rds file is now ready to be loaded in Escort Shiny app for further analysis in steps 2 and 3.

