---
title: "Generating embedding/trajectory objects for Escort R/Shiny app"
author: "Xiaoru Dong and Rhonda Bacher"
output: html_document
fontsize: 15pt
vignette: >
  %\VignetteIndexEntry{Escort}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

This tutorial demonstrates how to create an object that stores embedding and trajectory information that can then be used in the Shiny app directly in Steps 2 and 3 (bypassing Step 1). This also allows users to evaluate embeddings in batches with ease.

If you have not yet installed Escort, please revisit the main ["Running Escort in R"](https://www.rhondabacher.com/docs-escort/vignettes.html) vignette.

We will create the Shiny app-ready object using the function `prepTraj` provided in the Escort package. It requires three arguments: a reduced dimension matrix, estimated pseudotime, and the fitted trajectory.

First, we will make use of the sample dataset provided in the package.

### Load Escort and sample data
```{r, eval=FALSE}
library(Escort)
data("step0_clean_dyntoy_L3")
df <- Escort::HVGs_quick(normcounts = norm_counts)

```

For this example, we generate an embedding that consists of using 2,000 highly variable genes and using PCA as the dimension reduction.

### Generate embedding
```{r, eval=FALSE}
# get highly variable genes
gene.var <- quick_model_gene_var(norm_counts)
genes.HVGs <- rownames(gene.var)[1:2000]
# use PCA to reduce the dimensionality
embedding1 <- getDR_2D(norm_counts[genes.HVGs,], "PCA")
```

The object `embedding1` represents the `dimred` argument in the `prepTraj` function. If using alternative procedures, make sure that your `dimred` object is a data frame with two columns representing the 2D embedding coordinates, with each row representing a cell. The row names of `dimred` should be unique and accurately reflect cell identifiers.

We also need to go ahead and fit a trajectory to ensure the embedding works well not just independently, but also in the context of a specific method (Step 3 evaluation). Here we will proceed with Slingshot, however, other methods can be utilized.

```{r, eval=FALSE}
library(slingshot)
library(mclust)
cls1 <- Mclust(embedding1)$classification
ti_out1 <- slingshot(data=embedding1, clusterLabels=cls1)
rawpse1 <- slingPseudotime(ti_out1, na=T)
ls_fitLine1 <- lapply(slingCurves(ti_out1), function(x) x$s[x$ord,])
```

The object `rawpse1` will be input to the `PT` argument in the `prepTraj` function. Note that this object should be either a data frame or a vector. In the case of a data frame, each column should represent a lineage with pseudotime values, and rows should align with cell names in `dimred` object.

The object `ls_fitLine1` will be input to the `fitLine` argument in the `prepTraj` function. Note that this object represents the trajectory with line segments between pairs of points. It is formatted as a data frame, where columns "x0", "y0", "x1", and "y1" denote the starting and ending coordinates of each line segment.


### Prepare the embedding object

Now we use the `prepTraj` function to create an object for evaluation and save it as an .rds file.

```{r, eval=FALSE}
embed_obj <- prepTraj(dimred=embedding1, PT=rawpse1, fitLine=ls_fitLine1)
filepath <- "path/to/your_file.rds" # Specify your file path here.
saveRDS(embed_obj, file = filepath)
```

This .rds file is now ready to be loaded in Escort Shiny app for further analysis in Steps 2 and 3!

Note that you'll also need to upload the normalized data as an .rds object:

```{r, eval=FALSE}
filepath <- "path/to/file/norm_counts.rds" # Specify your file path here.
saveRDS(norm_counts, file = filepath)
```



### Generating multiple embedding objects

Now if you want to generate multiple embeddings to compare, here's how to go about this.

Say we want to evaluate different choices of highly variable genes, we'll write a function and a loop to generate and save these quickly.

```{r, eval=FALSE}
myembeds <- function(varyg) {
  genes.HVGs <- rownames(gene.var)[1:varyg]
  embedding1 <- getDR_2D(norm_counts[genes.HVGs,], "PCA")
  cls1 <- Mclust(embedding1)$classification
  ti_out1 <- slingshot(data=embedding1, clusterLabels=cls1)
  rawpse1 <- slingPseudotime(ti_out1, na=T)
  ls_fitLine1 <- lapply(slingCurves(ti_out1), function(x) x$s[x$ord,])
  
  embed_obj <- prepTraj(dimred=embedding1, PT=rawpse1, fitLine=ls_fitLine1)
  filepath <- paste0("path/to/embedding_HVG-", varyg, ".rds")
  saveRDS(embed_obj, file = filepath)
}

lapply(c(500, 1000, 1500, 2000, 5000), myembeds)
```


Now, we have generated 5 embeddings that we can upload directly into Step 2 & 3 for evaluation. In practice, you can vary many different processing choices including dimension reduction techniques and hyperparameters within slingshot (or the TI method of choice). 


